---
title: "Assignment 1"
author: "Ariz Kazani"
date: "2024-05-31"
output: pdf_document
---

# Assignment 1

## Name: Ariz Kazani

## Student ID: 101311311

------------------------------------------------------------------------

# Notes

```{r}
# TODO: add information about assignment and libraries used
#     - make sure to double check and double read each question
#     - format code ctr + shift + a
#     - test all code to make sure its still working

```

------------------------------------------------------------------------

# Solutions

------------------------------------------------------------------------

## Question 1. Function Creation and Vector Operations

### (a) Create a vector named sales that contains the following sales figures for a week: 250, 310, 450, 500, 620, 715, and 840.

```{r}
# vector that contains sales figures
sales <- c(250, 310, 450, 500, 620, 715, 840)
```

### (b) Write a function named **sales_summary** that takes a vector as input and returns the sum and mean of the vector. Test your function using the sales vector.

```{r}
sales_summary <- function(vector) {
  Sum <- sum(vector)
  Mean <- mean(vector)
  returnVar <- list(Sum = Sum, Mean = Mean)
  return(returnVar)
}

sales_summary(sales)
```

### (c) Write a function named **adjust_sales** that takes a vector and a percentage as inputs, adjusts each entry in the vector by the given percentage, and returns the adjusted vector in descending order. Test your function with the sales vector and a 10% increase.

```{r}
adjust_sales <- function(vector, percentage) {
  newVector <- vector * ((percentage + 100) / 100)
  sort(newVector, decreasing = TRUE)
  return(newVector)
}

adjust_sales(sales, 10)
```

### (d) Create another test for the **sales_summary** function with a random vector of 10 elements. Print the result to check if your function works correctly with different inputs.

```{r}
randomVector = c(22, 3, 24, 536, 774678, 895676, 57635, 24344, 123, 534)

sales_summary(randomVector)
```

### (e) Similarly, test the adjust_sales function with a random vector of 10 elements and a random percentage between 5% and 20%. Print the adjusted vector to ensure your function works correctly.

```{r}
adjust_sales(randomVector, 17)
```

### (f) Plot the original sales vector and the adjusted sales vector (from Part 3) on the same graph using different colors. Label the axes and add a legend.

```{r}
# TODO: fix this when the lecture comes out
# plot(
#   sales,
#   type = "o",
#   col = "blue",
#   ylim = range(c(sales, salesADJ)),
#   main = "Multiple Vectors Plot",
#   xlab = "Index",
#   ylab = "Value"
# )
# 
# lines(salesADJ, type = "o", col = "red")
# 
# legend(
#   "bottomright",
#   legend = c("Sales", "Adjusted Sales"),
#   col = c("blue", "red"),
#   lty = 1
# )
```

------------------------------------------------------------------------

## Question 2. Dataframe Operations and Descriptive Statistics

### (a) Create a dataframe named students with the following data:

-   Name: "Alice", "Bob", "Charlie", "David", "Eva"
-   Age: 23, 22, 24, 21, 23
-   Score: 85, 92, 78, 88, 90

```{r}
students = data.frame(
  Name = c("Alice", "Bob", "Charlie", "David", "Eva"),
  Age = c(23, 22, 24, 21, 23),
  Score = c(85, 92, 78, 88, 90)
)
```

### (b) Add a new column to the students dataframe named Passed with a value of TRUE if the Score is 80 or above, and FALSE otherwise.

```{r}
students$Passed <- students$Score >= 80

students
```

### (c) Calculate the mean, median, and standard deviation of the Age and Score columns in the students dataframe.

```{r}
meanAge = mean(students$Age)
medianAge = median(students$Age)
stdAge = sd(students$Age)
meanScore = mean(students$Score)
medianScore = median(students$Score)
stdScore = sd(students$Score)
```

### (d) Identify the student(s) with the highest score and display their details.

```{r}
maxScore = max(students$Score)

studentsWMS <- students[students$Score == maxScore, ]

studentsWMS
```

### (e) Filter the dataframe to show only the students who passed and save it as a new dataframe named passed_students.

```{r}
passed_students <- students[students$Passed, ]
```

### (f) Create a bar chart showing the scores of all students. Use different colors for those who passed and those who did not.

```{r}
# TODO: fill this out
```

### (g) Write a short summary (3-5 sentences) interpreting the statistical results and the bar chart created in the previous steps.

```{r}
# TODO: fill this out
```

------------------------------------------------------------------------

## 3. Advanced Data Manipulation and Visualization

### (a) Create a dataframe named employees with the following data:

-   EmployeeID: 101, 102, 103, 104, 105
-   Name: "John", "Jane", "Doe", "Smith", "Emily"
-   Department: "Sales", "HR", "IT", "Finance", "Marketing"
-   Salary: 60000, 65000, 70000, 72000, 68000
-   Experience: 3, 7, 5, 10, 4

```{r}
employees <- data.frame(
  EmployeeID = c(101, 102, 103, 104, 105),
  Name = c("John", "Jane", "Doe", "Smith", "Emily"),
  Department = c("Sales", "HR", "IT", "Finance", "Marketing"),
  Salary = c(60000, 65000, 70000, 72000, 68000),
  Experience = c(3, 7, 5, 10, 4)
)

employees
```

### (b) Calculate the mean and median salary for each department. Write a function named department_summary that returns a summary dataframe containing the department name, mean salary, and median salary.

```{r}
department_summary <- function(dataFrame) {
  getMeanMedian <- function(df, name) {
    dfWithName <- df[df$Department == name, ]
    returnDF = data.frame(
      DepartmentName = name,
      MeanSalary = mean(dfWithName$Salary),
      MedianSalary = median(dfWithName$Salary)
    )
    return(returnDF)
  }
  
  departments <- unique(dataFrame$Department)
  df = data.frame()
  for (dep in departments) {
    df <- rbind(df, getMeanMedian(dataFrame, dep))
  }
  return(df)
}

department_summary(employees)
```

### (c) Identify and display details of the employee with the highest salary in each department. Write a function named top_earner to achieve this.

```{r}
top_earner <- function(dataFrame) {
  getMax <- function(df, dep) {
    peopleByDep <- df[df$Department == dep, ]
    maxSal <- max(peopleByDep$Salary)
    topEarner <- peopleByDep[peopleByDep$Salary == maxSal, ]
    return(topEarner)
  }
  departments <- unique(dataFrame$Department)
  df = data.frame()
  for (dep in departments) {
    df <- rbind(df, getMax(dataFrame, dep))
  }
  return(df)
}

top_earner(employees)
```

### (d) Add a new column to the employees dataframe named AdjustedSalary, which is the Salary adjusted for experience (increase by 2% for each year of experience).

```{r}
employees$AdjustedSalary <- employees$Salary *
  ((100 + employees$Experience * 2) / 100)
```

### (e) Filter the dataframe to show only employees with an adjusted salary above 70,000 and save it as a new dataframe named high_earners.

```{r}
high_earners <- employees[employees$AdjustedSalary > 70000, ]
```

### (f) Create a boxplot to compare the distribution of original salaries and adjusted salaries across different departments. Add appropriate labels and a title.

```{r}
# TODO: fill this out
```

### (g) Write a short analysis (4-6 sentences) interpreting the results from the summary statistics, top earners, and the boxplot.

```{r}
# TODO: fill this out
```

------------------------------------------------------------------------

## 4. Exploring Dataframes with Multiple Operations

### (a) Create a dataframe named products with the following data:

-   ProductID: 201, 202, 203, 204, 205
-   ProductName: "Laptop", "Smartphone", "Tablet", "Headphones", "Smartwatch"
-   Category: "Electronics", "Electronics", "Electronics", "Accessories", "Electronics"
-   Price: 1200, 800, 600, 200, 350
-   QuantitySold: 150, 200, 300, 400, 250

```{r}
products <- data.frame(
  ProductID = c(201, 202, 203, 204, 205),
  ProductName =  c("Laptop", "Smartphone", "Tablet", "Headphones", "Smartwatch"),
  Category = c(
    "Electronics",
    "Electronics",
    "Electronics",
    "Accessories",
    "Electronics"
  ),
  Price = c(1200, 800, 600, 200, 350),
  QuantitySold = c(150, 200, 300, 400, 250)
)
```

### (b) Calculate the total revenue for each product (Price \* QuantitySold). Write a function named calculate_revenue that adds a new column Revenue to the products dataframe.

```{r}
calculate_revenue <- function(dataFrame) {
  dataFrame$Revenue <- dataFrame$Price * dataFrame$QuantitySold
  return(dataFrame)
}

products <- calculate_revenue(products)
```

### (c) Identify the product with the highest revenue and display its details.

```{r}
maxRevenue <- max(products$Revenue)
products[products$Revenue == maxRevenue, ]
```

### (d) Group the products by Category and calculate the total revenue for each category. Write a function named category_revenue that returns a summary dataframe with Category and TotalRevenue.

```{r}
category_revenue <- function(dataFrame) {
  dataByProduct <- function(dataFrame, category) {
    toteRev <- sum(dataFrame[dataFrame$Category == category, 'Revenue'])
    returnDf <- data.frame(Category = category, TotalRevenue = toteRev)
    return(returnDf)
  }
  catagories <- unique(dataFrame$Category)
  df <- data.frame()
  for (cat in catagories) {
    df <- rbind(df, dataByProduct(dataFrame, cat))
  }
  return(df)
}

category_revenue(products)
```

### (e) Create a bar chart to display the total revenue for each product, and use different colors for each category.

```{r}
# TODO: fill this out
```

### (f) Generate a scatter plot of Price versus QuantitySold with different colors for each category. Add a trend line to the plot.

```{r}
# TODO: fill this out
```

### (g) Write a detailed report (5-7 sentences) analyzing the

------------------------------------------------------------------------

## 5. Debugging Subsetting and Indexing Issues

Explain the issues with the code and provide the correct working code. Output the code to show that you have it corrected.

### (a)

```{r, eval=FALSE}
students <- data.frame(
  Name = c("Alice", "Bob", "Charlie", "David", "Eva"),
  Age = c(23, 22, 24, 21, 23),
  Score = c(85, 92, 78, 88, 90))
# Extracting ages of students who scored above 80
high_scorers_ages <- students[students$Score > 80][, "Age"]
print(high_scorers_ages)
```

##### Explanation:

`students[students$Score > 80]` is not correctly specifying rows and columns, to fix this we need to add a comma after specifying the columns like this `students[students$Score > 80,]`. Although it is not syntactically wrong we can also combine `students[students$Score > 80, ]` and `[, "Age"]` to look more clean `students[students$Score > 80, "Age"]`.

##### Corrected Code:

```{r}
students <- data.frame(
  Name = c("Alice", "Bob", "Charlie", "David", "Eva"),
  Age = c(23, 22, 24, 21, 23),
  Score = c(85, 92, 78, 88, 90))
# Extracting ages of students who scored above 80
high_scorers_ages <- students[students$Score > 80, "Age"]
print(high_scorers_ages)
```

### (b)

```{r, eval=FALSE}
employee_list <- list(
  Name = "John",
  Age = 30,
  Department = "HR",
  Salary = 50000
)
# Accessing the salary of the employee
salary <- employee_list["Salaries"]
print(salary)
```

##### Explanation:

The salary of the employee is not correctly being accessed. It was assigned as `Salary` but we are attempting to get `"Salaries"`. Instead we need to get `"Salary"`.

##### Corrected Code:

```{r}
employee_list <- list(
  Name = "John",
  Age = 30,
  Department = "HR",
  Salary = 50000
)
# Accessing the salary of the employee
salary <- employee_list["Salary"]
print(salary)
```

### (c)

```{r, eval=FALSE}
sales_data <- array(1:27, dim = c(3, 3, 3))
# Extracting the value in the second row, second column of the first matrix
value <- sales_data[3, 3, 0]
print(value)
```

##### Explanation:

The first issue is that we aren't correctly selecting the row and column if we want the second row and second column we want to get the data from `[2, 2,]`. the second issue is that we are not correctly accessing the first matrix, since R is 1 indexed and not 0, to get the first matrix we need to use 1 to access it and not 0: `sales_data[2, 2, 1]`.

##### Corrected Code:

```{r}
sales_data <- array(1:27, dim = c(3, 3, 3))
# Extracting the value in the second row, second column of the first matrix
value <- sales_data[2, 2, 1]
print(value)
```

### (d)

```{r, eval=FALSE}
products <- data.frame(
  ProductID = c(201, 202, 203, 204, 205),
  ProductName = c("Laptop", "Smartphone", "Tablet", "Headphones", "Smartwatch"),
  Category = c("Electronics", "Electronics", "Electronics", "Accessories", "Electronics"),
  Price = c(1200, 800, 600, 200, 350),
  QuantitySold = c(150, 200, 300, 400, 250)
)
# Extracting products with a price above 500
expensive_products <- products[products$Price >= "500", ]
print(expensive_products)
```

##### Explanation:

The first issue is that we are trying to compare string value with a numeric value. When this happens the numeric value gets converted into a string and gets compared lexicographically. This will not always give the desired result, for example `1200` gets converted into `"1200"` and then gets compared with `"500"`. Since `"1"` is smaller than `"5"` it will not add it. Another issue is that we want to get prices above 500 but we are also including 500 which is not what we want.

##### Corrected Code:

```{r}
products <- data.frame(
  ProductID = c(201, 202, 203, 204, 205),
  ProductName = c("Laptop", "Smartphone", "Tablet", "Headphones", "Smartwatch"),
  Category = c("Electronics", "Electronics", "Electronics", "Accessories", "Electronics"),
  Price = c(1200, 800, 600, 200, 350),
  QuantitySold = c(150, 200, 300, 400, 250)
)
# Extracting products with a price above 500
expensive_products <- products[products$Price > 500, ]
print(expensive_products)
```

------------------------------------------------------------------------

## 6. Analysis of the "trees" Dataset

This dataset has three variables (Girth, Height, Volume) on 31 felled black cherry trees.

### (a)

-   Load the "trees" dataset and check the structure with str().
-   Use apply() to return the mean values for the three variables (Girth, Height, Volume) and output these values.
-   Determine the number of trees with Volume greater than the mean Volume.

### (b)

-   Convert each Girth (diameter) to a radius $r$.
-   Calculate the cross-sectional area of each tree using $3.14 \times r$ ^2^.
-   Calculate and output the interquartile range (IQR) of the areas.

### (c)

-   Create a histogram of the areas calculated in part (b).
-   Title and label the axes.

### (d)

-   Identify the tree with the largest area.
-   Output its row number and the three measurements (Girth, Height, Volume) on one line

------------------------------------------------------------------------

## 7. Comprehensive Data Analysis and Function Creation

### (a)

-   Load the mtcars dataset.
-   Filter the dataset to include only cars with 6 or more cylinders and horsepower greater than 150. Save this filtered dataset as filtered_cars.

### (b)

-   Create a function named efficiency_score that calculates an efficiency score for each car based on the formula: $Efficiency Score$ = $\frac{mpg}{(hp \times wt)}$
-   Apply this function to the filtered_cars dataset and add the resulting scores as a new column named Efficiency.

### (c)

-   Identify rows where the Efficiency score is less than the 1st percentile or greater than the 99th percentile of all Efficiency scores.
-   Replace these outlier values with the mean Efficiency score of the remaining cars.

### (d)

-   Create a scatter plot of hp versus Efficiency, with points colored by the number of cylinders (cyl).
-   Add a trend line to the scatter plot.
-   Write a detailed analysis (6-8 sentences) interpreting the relationship between horsepower and efficiency, considering the number of cylinders.
