---
title: "Assignment 4"
author: "Ariz Kazani"
date: "`r Sys.Date()`"
output: pdf_document
---

# Assignment 4

## Name: Ariz Kazani

## Student ID: 101311311

------------------------------------------------------------------------

# Notes

```{r}
# Libraries

# NOTE: if you do not have any of the below libraries installed, 
# un-comment the line and run it

# install.packages("rmarkdown")
library(rmarkdown)

# install.packages("ggplot2")
library(ggplot2)

# useful libraries not required for this assignment
# install.packages("patchwork")
# library(patchwork)

# install.packages("dplyr")
# library(dplyr)

# install.packages("tidyr")
# library(tidyr)

# install.packages("gapminder")
# library(gapminder)

```

------------------------------------------------------------------------

# Solutions

------------------------------------------------------------------------

## 1. Discrete Probabilistic Modeling - Markov Chains

### A. Define a simple Markov Chain with 3 states: A, B, and C. Create a transition matrix where the probability of transitioning from any state to another is as follows:

-   P(A -\> B) = 0.3, P(A -\> C) = 0.7
-   P(B -\> A) = 0.6, P(B -\> C) = 0.4
-   P(C -\> A) = 0.8, P(C -\> B) = 0.2

```{r}

pTrans <- matrix(c(
  0, 0.6, 0.8,
  0.3, 0, 0.2,
  0.7, 0.4, 0
), nrow = 3, byrow = TRUE)

pTrans
```

### B. Write an R function to simulate the Markov Chain for 100 steps starting from state A. Plot the state

transitions over time.

```{r}

```

### C. Compute the stationary distribution of the Markov Chain using the transition matrix.

```{r}

```

### D. Verify the stationary distribution result by simulating the chain for a large number of steps (e.g., 10,000) and comparing the proportion of time spent in each state with the stationary distribution.

```{r}

```

------------------------------------------------------------------------

## 2. Discrete Probabilistic Modeling - Basic Monte Carlo Simulation

### A. Implement a basic Monte Carlo simulation to estimate the value of pi. Use the following steps:

1.  Generate a large number of random points within a unit square.
2.  Count the number of points that fall inside the unit circle.
3.  Use the ratio of points inside the circle to the total number of points to estimate pi.

```{r}

```

### B. Write an R function to perform this simulation with 10,000 points.

```{r}

```

### C. Plot the generated points, differentiating between points inside and outside the circle. Overlay the circle on the plot.

```{r}

```

### D. Calculate the estimated value of pi from your simulation and compare it to the true value of pi. Discuss the accuracy of the estimate.

```{r}

```

------------------------------------------------------------------------

## 3. Linear Programming

### A. Formulate the following linear programming problem: Maximize Z = 3x~1~ + 2x~2~ Subject to:

-   2x~1~ + x~2~ \<= 20
-   4x~1~ − 5x~2~ \>= −10
-   x~1~ + 2x~2~ \<= 15
-   x~1~, x~2~ \>= 0

```{r}

```

### B. Solve the linear programming problem using the lpSolve package in R. Provide the R code and solution.

```{r}

```

### C. Interpret the solution, including the values of x1x_1x1 and x2x_2x2, and the maximum value of ZZZ.

```{r}

```

------------------------------------------------------------------------

## 4. Integer Programming

### A. Formulate the following integer programming problem: Maximize 𝑍 = 4𝑥1 + 3𝑥2 Subject to:

-   3x~1~ + 2x~2~ \<= 12
-   x~1~ − x~2~ \>= 1
-   x~1~, x~2~ \>= 0 and integers

```{r}

```

### B. Solve the integer programming problem using the ompr package in R. Provide the R code and solution.

```{r}

```

### C. Interpret the solution, including the values of x~1~ and x~2~, and the maximum value of Z.

```{r}

```

### D. Discuss the differences between the linear and integer programming solutions and the situations in which integer programming is necessary.

```{r}

```

------------------------------------------------------------------------

## 5. Bootstrap Methods

### A. Use the bootstrap method to estimate the mean and 95% confidence interval of the following dataset:

data \<- c(5.1, 4.9, 5.7, 5.5, 5.0, 4.8, 5.2, 5.6, 4.7, 5.3)

Write an R function to perform bootstrap resampling and calculate the mean for each resample. Perform 10,000 bootstrap resamples.

```{r}

```

### B. Construct a histogram of the bootstrap means and calculate the 95% confidence interval for the mean based on the bootstrap samples.

```{r}

```

### C. Compare the bootstrap confidence interval with the theoretical confidence interval based on the normal distribution. Discuss any differences.

```{r}

```

### D. Evaluate the impact of the sample size on the bootstrap confidence interval by repeating the bootstrap process with different sample sizes (e.g., n=5, n=20) and comparing the results.

```{r}

```

------------------------------------------------------------------------

## 6. Bootstrap Methods - Regression

### A. Consider the following dataset:

```         
set.seed(123)
x <- rnorm(100)
y <- 2 + 3*x + rnorm(100)
```

Fit a linear regression model 𝑦 = B~0~ + B~1~ x + e and use bootstrap resampling to estimate the standard errors of the regression coefficients.

```{r}

```

### B. Perform 10,000 bootstrap resamples and calculate the standard errors for B~0~ and B~1~.

```{r}

```

### C. Compare the bootstrap standard errors with the standard errors obtained from the original regression model.

```{r}

```

### D. Evaluate the impact of the sample size on the bootstrap standard errors by repeating the bootstrap process with different sample sizes (e.g., n=50, n=200) and comparing the results.

```{r}

```
